Reproducible development environments with rix

### 2023-12-23 paper.qmd
more on rix

```{r, echo = F}
library(rix)

rix(r_ver = "4.3.1",
    project_path = tempdir(),
    overwrite = TRUE,
    print = TRUE)
```

To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so
the first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in
this derivation. In this example, the package expressions are going to be pulled
from the `nixpkgs`

```{r}
library(rix)

rix(r_ver = "4.2.2",
    r_pkgs = c("dplyr", "chronicler"),
    project_path = tempdir(),
    overwrite = TRUE,
    print = TRUE)
```

>>>>

```
let
  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/976fa3369d722e76f37c77493d99829540d43845.tar.gz") {};
  system_packages = builtins.attrValues {
    inherit (pkgs) R ;
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
```

To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github respository, and in particular the commit starting
with `976fa336` is going to be used. It sould be noted that is thus possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependecnies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell
[R]{.proglang} will be started with the `--vanilla` flag (thus ignore any
startup options).

This expression can be saved in a file called `default.nix`, and the shell can
be built using the `nix-build` command. Once `nix-build` is done building the
development environment according to the specifications declared in the
`default.nix` file, the user can enter the development shell using `nix-shell`.
The user will thus be dropped into an interactive shell with all the packages
declared in the `default.nix` file and can now use this shell for work.

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite interesting. Because we thought this, we wrote the [rix]{.pkg}
which comes with the `rix()` function that helps generate such expressions quite
easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

```{r, eval = F}
library(rix)

rix(r_ver = "4.3.1",
    r_pkgs = c("dplyr", "chronicler"),
    project_path = tempdir(),
    overwrite = TRUE,
    print = TRUE)
```

Here, users do not need to look for the right commit of `nixpkgs`: it is enough
to specify the version of [R]{.proglang} they wish to use and the packages that
were current at the time will get used. But [rix]{.pkg} is quite flexible and
makes it easy to install specific versions of packages as well (which will get
installed from the CRAN archives) as well as packages hosted on Github and
\LaTeX packages as well:

```{r, eval = F}
rix(r_ver = "4.3.1",
    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
    system_pkgs = c("quarto", "git"),
    tex_pkgs = c(
          "amsmath",
          "framed",
          "fvextra",
          "environ",
          "fontawesome5",
          "orcidlink",
          "pdfcol",
          "tcolorbox",
          "tikzfill"
    ),
    git_pkgs = list(
                 list(package_name = "rix",
                   repo_url = "https://github.com/b-rodrigues/rix/",
                   branch_name = "master",
                   commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
                 list(package_name = "fusen",
                   repo_url = "https://github.com/ThinkR-open/fusen",
                   branch_name = "main",
                   commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
    ),
    ide = "rstudio",
    project_path = tempfile(),
    overwrite = TRUE)
```

The above call to `rix()` generates an expression that installs several
[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
older version of the package. If the user didn't not specify the version, the
version available as of the `nixpkgs` commit would be used), but also install
several TeXLive packages for authoring documents with \LaTeX, the development
versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
the RStudio integrated development environment.
